---
title: "Indistinguishable dyads: multilevel modelling"
subtitle: "Long format, lme4, random intercept per dyad"
---

The most common way to fit an APIM in R is with a multilevel model
in long format. The random intercept for `dyad_id` accounts for the
non-independence within dyads. Slopes are constrained to be equal
across dyad members by the data layout itself — the same variable
name appears for both members.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lme4)
library(lmerTest)
library(dplyr)

# Load the long-format data
load("../../../data/dyad_data.RData")

cat("Long format:", nrow(ddl), "rows,", ncol(ddl), "columns\n")
cat("Dyads:    ", length(unique(ddl$dyad_id)), "\n")
cat("Persons:  ", nrow(ddl), "\n")
```

## Step 1: estimate the ICC

Before any APIM, fit a null model and read the intraclass correlation.
This is the proportion of outcome variance that lives *between*
dyads.

```{r}
#| label: null-model

null_model <- lmer(satisfaction ~ 1 + (1 | dyad_id), data = ddl)
summary(null_model)
```

Look at the `summary()` output. The random-effects section reports two
variance components: the `dyad_id` variance ($\tau^2$, the between-dyad
variance in satisfaction intercepts) and the `Residual` variance
($\sigma^2$, the within-dyad variance). Together they partition the
total satisfaction variance into a part that is shared within each dyad
and a part that is unique to each individual. The fixed-effects intercept
(~4.92) is the estimated grand mean of satisfaction, pooled across all
200 individuals, before any predictors enter.

::: {.callout-tip}
## Reading the variance components
The `Groups` column shows the random-effect variance. `dyad_id`
is `tau^2`, the variance of the dyad-level random intercept.
`Residual` is `sigma^2`, the within-dyad residual variance.

The ICC is `tau^2 / (tau^2 + sigma^2)`. With our data, expect
something close to 0.25 — a quarter of satisfaction variance is
shared by the two members of each dyad.
:::

```{r}
#| label: icc

varcomp <- as.data.frame(VarCorr(null_model))
tau2  <- varcomp$vcov[varcomp$grp == "dyad_id"]
sigma2 <- varcomp$vcov[varcomp$grp == "Residual"]
icc   <- tau2 / (tau2 + sigma2)

cat("\nICC:", round(icc, 4), "\n")
cat("Interpretation: ", round(icc * 100, 1),
    "% of satisfaction variance is shared within dyads.\n", sep = "")
```

An ICC of this size is a clear signal that the dyad is the proper
unit of analysis. Standard OLS would underestimate standard errors
by roughly 15–20% on these data.

::: {.callout-note collapse="false"}
## What this ICC means in practice

An ICC of roughly 0.25 says that a quarter of the variation in
relationship satisfaction is driven by factors that the two partners
share — their household, their relationship history, their joint
financial circumstances. The remaining three-quarters is individual-level
variation: one partner's personal stressors, their own recovery habits,
their measurement error. If you had ignored the dyad structure and run a
standard OLS regression, the effective sample size would be roughly
$200 / (1 + 1 \times 0.25) \approx 160$ — you would have overstated your
precision by about 25% and inflated your Type I error rate. The random
intercept for `dyad_id` is the minimum fix.
:::

## Step 2: the full APIM

The full APIM in long format looks like a regression of satisfaction
on the actor's predictors, the partner's predictors, and any
dyad-level covariates, with a random intercept for dyad.

```{r}
#| label: apim

apim_mlm <- lmer(
  satisfaction ~ wnc + partner_wnc + recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(apim_mlm)
```

::: {.callout-tip}
## Reading the coefficients

Scan the fixed-effects table from the `summary()` output and compare each
coefficient to the known DGP value. The actor WNC effect (`wnc`) should
land near −0.30 — a one-unit increase in a person's own work–nonwork
conflict is associated with a 0.30-unit decrease in their own
relationship satisfaction, holding all else constant. The partner WNC
effect (`partner_wnc`) should land near −0.15 — half the magnitude of the
actor effect, but in the same direction. This is the partner *crossover*:
my partner's stress spills over and worsens my satisfaction, even after
controlling for my own stress. The actor recovery effect (`recovery`,
near +0.25) and partner recovery effect (`partner_recovery`, near +0.10)
tell the mirror story: recovery buffers satisfaction, and a partner's
recovery buffers mine too.

The two dyad-level covariates are interpretable as between-dyad
contrasts. `has_children` (near −0.10) means that, on average, couples
with children score about 0.10 points lower on satisfaction than couples
without, controlling for WNC and recovery. `dual_earner` (near +0.20)
means that dual-earner couples score higher — a compositional effect
consistent with higher resources buffering relationship strain. These
coefficients are estimated entirely from the 100 dyad-level observations,
so their standard errors are wider than the actor effects, which borrow
strength from 200 person-level observations.

The `lmerTest` package adds Satterthwaite-approximated degrees of freedom
and p-values to the `lme4` output. These p-values are approximate — do
not treat them as exact. For reporting, prefer the profile-likelihood
confidence intervals in Step 4; they are more accurate for mixed models,
especially for the variance components.

- `wnc` is the **actor** effect of WNC: how much the actor's own
  WNC affects their own satisfaction. The point estimate should be
  close to the DGP value of `−0.30`.
- `partner_wnc` is the **partner** effect: how much the partner's
  WNC affects the actor's satisfaction. The point estimate should
  be close to `−0.15`.
- `recovery` and `partner_recovery` follow the same pattern.
- `has_children` and `dual_earner` are dyad-level covariates with
  no partner split.
:::

The two predictors of substantive interest — WNC and recovery —
both show a clear actor effect and a smaller partner effect, in
the same direction. The actor effects are about twice the partner
effects, which is close to a [couple pattern](../../foundations/kpatterns.qmd)
with $k \approx 0.5$.

## Step 3: test whether the partner block is needed

The `anova()` output for `lme4` models reports the log-likelihood of each
model, the chi-square statistic ($\chi^2 = 2(\ell_{\text{full}} -
\ell_{\text{reduced}})$), the degrees of freedom (the number of
constrained parameters — here, both partner effects are constrained to
zero, so 2 df), and the p-value. A significant p-value means the partner
block is jointly needed: at least one of `partner_wnc` or
`partner_recovery` has a non-zero effect on satisfaction. A
non-significant p-value would mean the dyad is effectively an
actor-effects-only system, which would be a substantively important null
result — it would mean stress does not cross over between partners.

In our data, the LRT should be significant. The DGP has `p_wnc = −0.15`
and `p_rec = +0.10`, both non-zero. The test has adequate power at
N = 100 dyads to detect partner effects of these magnitudes.

Compare the full APIM to a reduced model that omits the partner
predictors. The likelihood ratio test tells you whether the partner
block is jointly useful.

```{r}
#| label: lrt-partner

actor_only <- lmer(
  satisfaction ~ wnc + recovery + has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

anova(actor_only, apim_mlm)
```


Profile-likelihood confidence intervals are the reporting standard for
mixed models because they account for the asymmetry of the likelihood
surface — Wald intervals assume the surface is quadratic, which is
reasonable for fixed effects but poor for variance components. Scan the
output for the intervals that exclude zero: `wnc`, `partner_wnc`,
`recovery`, and `partner_recovery` should all have confidence limits
that do not cross zero, confirming that both actor and partner effects
are reliably non-zero. The dyad-level covariates (`has_children`,
`dual_earner`) may have intervals that cross zero because they are
estimated from only 100 dyad-level observations. The random-intercept
standard deviation (`.sig01`) and residual standard deviation (`.sigma`)
are reported on the standard-deviation scale; their intervals should be
well above zero, confirming that both variance components are real.

## Step 4: profile-likelihood confidence intervals

Wald confidence intervals are faster to compute but less accurate
for variance components. Profile-likelihood intervals are the
standard for reporting.

```{r}
#| label: profile-ci

ci <- confint(apim_mlm, method = "profile")
print(round(ci, 4))
```

## What to take away

::: {.callout-takeaway}
## Key takeaways

- A random intercept for `dyad_id` is the minimum adjustment for
  dyadic non-independence. It corrects the standard errors and
  gives you the ICC for free.
- In long format, equal actor and partner slopes is the *default*,
  not a constraint you impose. The same variable name plays both
  roles.
- The partner effects in our simulated data are about half the
  size of the actor effects, with the same sign. This is a
  weak-couple pattern.
:::

## What to read next

- The [SEM wide tutorial](sem.qmd) fits the same model in `lavaan`
  with explicit equality constraints. You will see the same point
  estimates, plus a likelihood ratio test of the equality constraint.
- The [distinguishable MLM
  tutorial](../distinguishable/mlm-moderator.qmd) relaxes the
  equality of slopes and adds gender as a moderator.
- The [Hahn et al. (2014) replication
  tutorial](../moderated/hahn2014.qmd) adds the children ×
  partner-WNC moderation.
