---
title: "Indistinguishable dyads: SEM in wide format"
subtitle: "Wide format, lavaan, equality-constrained actor and partner slopes"
---

The multilevel model in the [previous tutorial](mlm.html) is the
workhorse for APIM work, but the SEM specification is more
transparent about *which* slopes are constrained equal, and it gives
you standard fit indices (CFI, TLI, RMSEA, SRMR) that you can
report alongside the parameter estimates. This tutorial fits the
indistinguishable APIM in `lavaan` with the Olsen & Kenny (2006)
specification.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lavaan)
library(dplyr)

load("../../../data/dyad_data.RData")

cat("Wide format:", nrow(ddw), "rows,", ncol(ddw), "columns\n")
```

## The model

In wide format, the two dyad members are represented by separate
columns with the suffixes `_a` and `_p`. The Olsen & Kenny (2006)
specification constrains the actor and partner slopes to be equal
by giving the two paths the same label.

```{r}
#| label: model-syntax

model_indist <- '
  # Actor (the "_a" member)
  satisfaction_a ~ a_wnc*wnc_a + p_wnc*wnc_p +
                   a_rec*recovery_a + p_rec*recovery_p +
                   c_child*has_children + c_dual*dual_earner

  # Partner (the "_p" member)
  satisfaction_p ~ a_wnc*wnc_p + p_wnc*wnc_a +
                   a_rec*recovery_p + p_rec*recovery_a +
                   c_child*has_children + c_dual*dual_earner

  # Equal intercepts
  satisfaction_a ~ alpha*1
  satisfaction_p ~ alpha*1

  # Equal residual variances
  satisfaction_a ~~ sigma2*satisfaction_a
  satisfaction_p ~~ sigma2*satisfaction_p

  # Predictor variances and covariances
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p

  # Residual covariance (the dyad-level non-independence)
  satisfaction_a ~~ res_cov*satisfaction_p
'
```

::: {.callout-tip}
## Reading the labels

- `a_wnc` is used for *both* the path `wnc_a → satisfaction_a` (the
  actor effect on the actor's own satisfaction) and the path
  `wnc_p → satisfaction_p` (the actor effect on the partner's
  satisfaction). Same label = constrained equal.
- `p_wnc` is used for the partner effect. The constraint forces
  `a_wnc = p_wnc` to be tested separately — the [distinguishable
  SEM wide tutorial](../distinguishable/sem-wide.html) shows how.
- `alpha`, `sigma2`, and `res_cov` are shared labels for the
  intercepts, residual variances, and residual covariance.
:::

## Fit the model

```{r}
#| label: fit

fit_indist <- sem(model_indist, data = ddw)

summary(fit_indist, standardized = TRUE, fit.measures = TRUE)
```

## Compare to the multilevel model

The point estimates should match the [MLM tutorial](mlm.html) to
three or four decimal places. The differences are in the standard
errors (Wald vs. profile-likelihood) and in the inferential
machinery (likelihood ratio tests in `lavaan` vs. in `lme4`).

```{r}
#| label: compare
#| message: false
#| warning: false

library(lme4)
library(lmerTest)

apim_mlm <- lmer(
  satisfaction ~ wnc + partner_wnc + recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

cat("--- MLM coefficients ---\n")
print(round(fixef(apim_mlm), 4))

cat("\n--- SEM coefficients (regression) ---\n")
pe <- parameterEstimates(fit_indist, standardized = TRUE)
pe[pe$op == "~", c("lhs", "rhs", "est", "se", "pvalue")]
```

The two sets of estimates should be nearly identical.

## Fit indices

```{r}
#| label: fit-indices

fits <- c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "srmr")
fitMeasures(fit_indist, fits)
```

The chi-square test of exact fit is a strict test; with N = 100
dyads and a model that is correctly specified for the data, you
should still see a significant chi-square — this is a well-known
limitation of the chi-square test for SEM. The other indices
(CFI, TLI, RMSEA, SRMR) are alternative fit criteria that are less
sensitive to sample size.

For an *indistinguishable* model on dyadic data, the SRMR is the
most informative index. Values below 0.08 are conventionally
considered acceptable. (See the [Hu & Bentler (1999) cutoffs
discussion](../../references/index.html) for context.)

## Test the equality constraint

The most informative test for indistinguishable dyads is whether
the equality constraint fits as well as the unconstrained model.
Fit the unconstrained model and compare.

```{r}
#| label: lrt-indistinguishability

model_unconstrained <- '
  satisfaction_a ~ a_wnc_h*wnc_a + p_wnc_h*wnc_p +
                   a_rec_h*recovery_a + p_rec_h*recovery_p +
                   c_child_h*has_children + c_dual_h*dual_earner
  satisfaction_p ~ a_wnc_w*wnc_p + p_wnc_w*wnc_a +
                   a_rec_w*recovery_p + p_rec_w*recovery_a +
                   c_child_w*has_children + c_dual_w*dual_earner
  satisfaction_a ~ int_a*1
  satisfaction_p ~ int_p*1
  satisfaction_a ~~ var_a*satisfaction_a
  satisfaction_p ~~ var_p*satisfaction_p
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p
  satisfaction_a ~~ res_cov*satisfaction_p
'

fit_unconstrained <- sem(model_unconstrained, data = ddw)

cat("--- LRT: unconstrained vs indistinguishable ---\n")
lavTestLRT(fit_unconstrained, fit_indist)
```

A non-significant *p*-value supports indistinguishability. In our
data, the equality constraint is tenable, which is consistent with
the [data-generating process](../../data/dyad-data.html) (equal
slopes across gender).

## What to take away

::: {.callout-takeaway}
## Key takeaways

- The wide-format SEM gives you the same point estimates as the
  MLM, with explicit labels for which slopes are constrained equal.
- The Olsen & Kenny (2006) specification is the standard reference
  for this model.
- The likelihood ratio test against the unconstrained model is the
  gold standard for testing distinguishability.
:::

## What to read next

- The [distinguishable SEM wide
  tutorial](../distinguishable/sem-wide.html) generalises this
  specification to the three-model sequence (unconstrained →
  slopes-equal → fully constrained) and adds the k-pattern tests.
- The [SEM with moderation
  tutorial](../distinguishable/sem-moderation.html) shows how to
  add gender × predictor interactions at the within-dyad level.
