---
title: "Distinguishable dyads: MLM moderator"
subtitle: "Long format, lme4, gender × predictor interactions"
---

The simplest way to add gender to the APIM is to fit a multilevel
model in long format with `gender` as a factor that interacts with
every actor and partner predictor. This is the "version B" approach
in the original workshop materials — it decomposes actor and
partner pathways properly.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lme4)
library(lmerTest)
library(dplyr)
library(interactions)

load("../../../data/dyad_data.RData")

cat("Dyads:    ", length(unique(ddl$dyad_id)), "\n")
cat("Persons:  ", nrow(ddl), "\n\n")
cat("Gender distribution:\n")
print(table(ddl$gender))
```

## Step 1: the baseline model (no gender)

The baseline is the same as the [indistinguishable
tutorial](../indistinguishable/mlm.html) APIM. It estimates pooled
slopes across gender.

```{r}
#| label: baseline

model_base <- lmer(
  satisfaction ~ wnc + partner_wnc + recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(model_base)
```

## Step 2: add gender × predictor interactions

The full moderator model interacts `gender` with every actor and
partner predictor. The two dyad-level covariates (`has_children`,
`dual_earner`) stay as main effects — they have no within-dyad
variance to moderate.

```{r}
#| label: moderator

model_moderator <- lmer(
  satisfaction ~ wnc * gender + partner_wnc * gender +
    recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(model_moderator)
```

::: {.callout-warning}
## Why "version B"?
The simpler formulation `wnc + partner_wnc + gender` (without the
`*` interactions) would conflate the actor and partner pathways
and is **not** recommended. Always use the decomposed form above:
`(wnc + partner_wnc) * gender`. This gives you separate gender
effects for the actor and partner pathways.
:::

## Step 3: likelihood ratio test

The likelihood ratio test of the baseline vs the moderator model
tests the joint null that all four gender × predictor interactions
are zero.

```{r}
#| label: lrt

lrt <- anova(model_base, model_moderator)
print(lrt)
```

In our data, the *p*-value should be large — the data-generating
process has equal slopes across gender. The LRT confirms what we
already know from the DGP.

## Step 4: simple slopes

If the LRT *had* been significant, the next step would be to extract
simple slopes for the actor and partner effects separately for each
gender. The `interactions` package has a clean interface for this.

```{r}
#| label: simple-slopes

sim_awnc <- sim_slopes(model_moderator, pred = wnc,
                       modx = gender, johnson_neyman = FALSE)
print(sim_awnc)
```

The `sim_slopes` function returns the slope of `wnc` on satisfaction
at each level of `gender` (male, female), with standard errors and
confidence intervals. In our data, the two slopes are nearly equal —
the gender × wnc interaction is small and not statistically
significant.

## Step 5: profile-likelihood confidence intervals

```{r}
#| label: profile-ci

ci <- confint(model_moderator, method = "profile")
print(round(ci, 4))
```

## Interpreting the intercept

`Male` is the reference group in R's default contrast coding. The
fixed-effect intercept is the male intercept. The `genderfemale`
coefficient is the *female minus male* difference in intercepts. In
our data, this difference should be close to the DGP value of
$-0.15$ ($\alpha_f - \alpha_m = 4.85 - 5.00$).

::: {.callout-tip}
## Reading the gender × predictor interaction

- `wnc:genderfemale` is the difference in the actor effect of WNC
  between females and males. Zero in our DGP.
- `partner_wnc:genderfemale` is the difference in the partner
  effect of WNC between females and males. Zero in our DGP.

If you fit this model on real data with non-zero gender moderation,
the corresponding interaction will be significantly different from
zero, and `sim_slopes` will return different slopes for males and
females.
:::

## What to take away

::: {.callout-takeaway}
## Key takeaways

- The MLM moderator approach fits a single long-format model with
  gender × predictor interactions. It is the most common
  specification in applied APIM work.
- Always use the decomposed form `(wnc + partner_wnc) * gender` —
  the simpler form conflates actor and partner pathways.
- The LRT of the moderator vs the baseline tests whether gender
  moderates any of the actor or partner effects *jointly*.
- Simple slopes are the recommended way to report the conditional
  actor and partner effects at each level of gender.
:::

## What to read next

- The [SEM with moderation tutorial](sem-moderation.html) shows
  the same specification in `lavaan` with calculated simple
  slopes.
- The [SEM wide tutorial](sem-wide.html) gives you the Olsen &
  Kenny (2006) three-model sequence and the k-pattern tests.
- The [two-intercept tutorial](two-intercept.html) shows a
  parameterisation trick that estimates both intercepts directly.
