---
title: "Distinguishable dyads: SEM in wide format"
subtitle: "The three-model sequence and the k-pattern tests"
---

The wide-format SEM is the classical specification for the
distinguishable APIM. The three nested models — unconstrained,
slopes-equal, fully constrained — give you three likelihood ratio
tests, each of which has a clean substantive interpretation.

This tutorial reproduces the Olsen & Kenny (2006) specification
and the Kenny & Ledermann (2010) k-pattern tests.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lavaan)
library(dplyr)

load("../../../data/dyad_data.RData")

cat("Wide format:", nrow(ddw), "rows\n")
```

## Model 1: unconstrained (all paths free)

The unconstrained model estimates a separate slope for every path.
Actor (male) and partner (female) intercepts, slopes, and residual
variances are all free to differ.

```{r}
#| label: model-unconstrained

model_unconstrained <- '
  satisfaction_a ~ a_h*wnc_a + p_h*wnc_p + ar_h*recovery_a +
                   pr_h*recovery_p + c_h*has_children + d_h*dual_earner
  satisfaction_p ~ a_w*wnc_p + p_w*wnc_a + ar_w*recovery_p +
                   pr_w*recovery_a + c_w*has_children + d_w*dual_earner
  satisfaction_a ~ int_a*1
  satisfaction_p ~ int_p*1
  satisfaction_a ~~ var_a*satisfaction_a
  satisfaction_p ~~ var_p*satisfaction_p
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p
  satisfaction_a ~~ res_cov*satisfaction_p
'

fit_unconstrained <- sem(model_unconstrained, data = ddw)
summary(fit_unconstrained, standardized = TRUE, fit.measures = TRUE)
```

## Model 2: slopes equal, intercepts free

The slopes-equal model constrains actor and partner slopes to be
the same. Intercepts and residual variances are free to differ.

```{r}
#| label: model-slopes-equal

model_slopes_equal <- '
  satisfaction_a ~ a*wnc_a + p*wnc_p + ar*recovery_a +
                   pr*recovery_p + c*has_children + d*dual_earner
  satisfaction_p ~ a*wnc_p + p*wnc_a + ar*recovery_p +
                   pr*recovery_a + c*has_children + d*dual_earner
  satisfaction_a ~ int_a*1
  satisfaction_p ~ int_p*1
  satisfaction_a ~~ var_h*satisfaction_a
  satisfaction_p ~~ var_w*satisfaction_p
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p
  satisfaction_a ~~ res_cov*satisfaction_p
'

fit_slopes_equal <- sem(model_slopes_equal, data = ddw)
summary(fit_slopes_equal, standardized = TRUE, fit.measures = TRUE)
```

## Model 3: fully constrained (slopes, intercepts, variances equal)

The fully-constrained model forces every parameter to be equal
across the two roles. This is the indistinguishable model.

```{r}
#| label: model-full-equal

model_full_equal <- '
  satisfaction_a ~ a*wnc_a + p*wnc_p + ar*recovery_a +
                   pr*recovery_p + c*has_children + d*dual_earner
  satisfaction_p ~ a*wnc_p + p*wnc_a + ar*recovery_p +
                   pr*recovery_a + c*has_children + d*dual_earner
  satisfaction_a ~ alpha*1
  satisfaction_p ~ alpha*1
  satisfaction_a ~~ var_res*satisfaction_a
  satisfaction_p ~~ var_res*satisfaction_p
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p
  satisfaction_a ~~ res_cov*satisfaction_p
'

fit_full_equal <- sem(model_full_equal, data = ddw)
summary(fit_full_equal, standardized = TRUE, fit.measures = TRUE)
```

## The three nested LRTs

```{r}
#| label: lrts

cat("LRT 1: Unconstrained vs Slopes Equal\n")
cat("(Do slopes differ by gender?)\n\n")
lrt1 <- lavTestLRT(fit_unconstrained, fit_slopes_equal)
print(lrt1)

cat("\nLRT 2: Slopes Equal vs Fully Constrained\n")
cat("(Do intercepts differ by gender?)\n\n")
lrt2 <- lavTestLRT(fit_slopes_equal, fit_full_equal)
print(lrt2)

cat("\nLRT 3: Unconstrained vs Fully Constrained\n")
cat("(Overall indistinguishability)\n\n")
lrt3 <- lavTestLRT(fit_unconstrained, fit_full_equal)
print(lrt3)
```

::: {.callout-tip}
## Reading the three tests

- **LRT 1** tests whether actor and partner slopes are equal. A
  non-significant *p*-value supports the *couple pattern* (equal
  slopes). Significant *p*-value means the dyads are
  distinguishable by effects, not just by means.
- **LRT 2** tests whether the intercepts and residual variances
  are equal, given equal slopes. A significant *p*-value means
  the dyads are distinguishable by *means*.
- **LRT 3** tests overall indistinguishability (slopes, intercepts,
  residual variances all equal). A non-significant *p*-value
  means the dyads are indistinguishable on every parameter.

In our data, LRT 1 should be non-significant (equal slopes) and
LRT 2 should be significant (different intercepts). This is the
standard empirical pattern: distinguishable by means, not by
effects.
:::

## Fit indices comparison

Build a side-by-side table of fit measures for the three models.

```{r}
#| label: fit-indices

fits <- c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "srmr")

fit_indices <- data.frame(
  unconstrained   = fitMeasures(fit_unconstrained,   fits),
  slopes_equal    = fitMeasures(fit_slopes_equal,    fits),
  full_equal      = fitMeasures(fit_full_equal,      fits)
)

print(round(fit_indices, 4))
```

## What to take away

::: {.callout-takeaway}
## Key takeaways

- The wide-format SEM gives you the three classical tests of
  distinguishability: slopes, intercepts, overall.
- LRT 1 (slopes) is the most important test for the APIM: it
  tells you whether the *effects* of predictors differ by role.
- LRT 2 (intercepts) tells you whether the dyads are
  distinguishable by *means* (often they are, even when slopes
  are equal).
- LRT 3 (overall) is the strict test; it is rarely used in
  practice because LRT 1 and LRT 2 are more informative.
:::

## What to read next

- The [two-intercept tutorial](two-intercept.html) — recommended
  when LRT 1 is non-significant but LRT 2 is significant (i.e.
  dyads are distinguishable by means, not effects).
- The [Hahn et al. (2014) replication
  tutorial](../moderated/hahn2014.html) — moderated APIM with a
  dyad-level moderator.
