---
title: "Distinguishable dyads: SEM with moderation"
subtitle: "Two-level SEM with within-dyad gender interactions"
---

The [MLM moderator tutorial](mlm-moderator.html) added gender ×
predictor interactions to a long-format multilevel model. This
tutorial does the same thing in `lavaan`, using the two-level SEM
specification that decomposes variance into within-dyad and
between-dyad components.

The two specifications should give you nearly identical estimates
for the within-dyad effects. The SEM version is more transparent
about which interactions are at the within level and which
covariates are at the between level.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lavaan)
library(dplyr)

load("../../../data/dyad_data.RData")

# Numeric gender indicator for interactions
# Male = 1, Female = 2 (matches R's default factor coding)
ddl$gender_num <- as.numeric(ddl$gender)

# Manual interaction columns
ddl$wnc_x_gender        <- ddl$wnc * ddl$gender_num
ddl$partner_wnc_x_gender <- ddl$partner_wnc * ddl$gender_num

cat("Long format:", nrow(ddl), "rows\n")
cat("Gender distribution:\n")
print(table(ddl$gender))
```

## Baseline: two-level SEM without gender moderation

Place the four within-dyad predictors at the within level and the
two dyad-level covariates at the between level. The two-level SEM
is the SEM analogue of the [MLM tutorial](mlm-moderator.html).

```{r}
#| label: baseline-2level

model_base <- '
  level: within
    satisfaction ~ a_w*wnc + p_w*partner_wnc + ar_w*recovery +
      pr_w*partner_recovery

  level: between
    satisfaction ~ c_child*has_children + c_dual*dual_earner
'

fit_base <- sem(model_base, data = ddl, cluster = "dyad_id")
summary(fit_base, standardized = TRUE, fit.measures = TRUE)
```

## Add gender × predictor interactions

The within-dyad level now includes the gender main effect and the
gender × actor/partner WNC interactions. The between-dyad level
is unchanged.

```{r}
#| label: moderator-2level

model_moderator <- '
  level: within
    satisfaction ~ a_w*wnc + p_w*partner_wnc + ar_w*recovery +
      pr_w*partner_recovery +
      g_w*gender_num +
      aw_int*wnc_x_gender + pw_int*partner_wnc_x_gender

  level: between
    satisfaction ~ c_child*has_children + c_dual*dual_earner

  # ── Simple slopes ──────────────────────────────
  # gender_num = 1 (male), gender_num = 2 (female)

  # Actor WNC → satisfaction
  ss_wnc_M  := a_w + aw_int * 1
  ss_wnc_F  := a_w + aw_int * 2

  # Partner WNC → satisfaction
  ss_pwnc_M := p_w + pw_int * 1
  ss_pwnc_F := p_w + pw_int * 2
'

fit_moderator <- sem(model_moderator, data = ddl, cluster = "dyad_id")
summary(fit_moderator, standardized = TRUE, fit.measures = TRUE)
```

## Reading the calculated simple slopes

The `:=` operator in `lavaan` defines a user-calculated parameter.
The simple slopes are functions of the base coefficient and the
interaction coefficient, evaluated at the male and female values
of `gender_num`.

```{r}
#| label: simple-slopes

pe <- parameterEstimates(fit_moderator)
pe[grepl("^ss_", pe$label), c("label", "est", "se", "pvalue")]
```

::: {.callout-tip}
## Comparing to the MLM estimates

The within-dyad SEM `aw_int` coefficient (gender × actor WNC) and
`pw_int` (gender × partner WNC) should match the corresponding
`wnc:genderfemale` and `partner_wnc:genderfemale` interactions in
the [MLM moderator tutorial](mlm-moderator.html#step-3-likelihood-ratio-test).
The simple slopes for males and females should also match the
`sim_slopes()` output.
:::

## Why "model is saturated" matters here

The two-level SEM with the within-dyad interactions has just-identified
within-level equations: the four predictors plus three interaction
terms and a gender main effect consume all the within-dyad degrees of
freedom. This means:

- The chi-square test of model fit is exactly zero (or near it)
  by construction.
- The standard chi-square difference test is *not* available for
  comparing the baseline and moderator models.

If you want a formal LRT, the right move is to fit both models in
the wide format (see the [SEM wide tutorial](sem-wide.html)), or to
use the [MLM moderator approach](mlm-moderator.html#step-3-likelihood-ratio-test),
where the LRT is well-defined.

## What to take away

::: {.callout-takeaway}
## Key takeaways

- The two-level SEM makes the within/between decomposition of
  effects explicit.
- Gender × predictor interactions live at the within-dyad level.
- Dyad-level covariates (no within-dyad variance) live at the
  between-dyad level.
- Calculated simple slopes via `:=` give you the conditional
  effects at each level of the moderator.
- The model is just-identified within-dyad, so the chi-square LRT
  is not available; use the MLM version for hypothesis testing.
:::

## What to read next

- The [SEM wide tutorial](sem-wide.html) — wide format, the
  three-model sequence, the k-pattern tests.
- The [two-intercept tutorial](two-intercept.html) — a
  parameterisation trick that estimates both intercepts directly.
- The [Hahn et al. (2014) replication
  tutorial](../moderated/hahn2014.html) — moderated APIM with a
  dyad-level moderator.
