---
title: "Two-intercept models"
subtitle: "Estimating both intercepts directly with the 0 + gender trick"
---

When dyads are distinguishable by *means* but not by *effects* —
the common empirical pattern — you usually want to report the
absolute mean of each group, not the contrast with a reference.
The two-intercept model is a parameterisation trick that estimates
both intercepts directly.

The trick: in R's formula syntax, `0 + gender` suppresses the
global intercept and replaces it with one intercept per level of
`gender`. You get two intercepts, no reference group, and a clean
likelihood ratio test for the "equal intercepts" null.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lme4)
library(lmerTest)
library(lavaan)
library(dplyr)

load("../../../data/dyad_data.RData")
```

## The three nested models

The two-intercept approach compares three nested models. The
ordering is: most constrained → recommended baseline → full model.

### Model 1: equal intercepts, equal slopes

The most constrained model. Single intercept, slopes pooled.

```{r}
#| label: model1

model1_mlm <- lmer(
  satisfaction ~ wnc + partner_wnc + recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(model1_mlm)
```

### Model 2: different intercepts, equal slopes (recommended)

The recommended baseline. Two intercepts, slopes pooled. Note the
`0 + gender` syntax — this replaces the global intercept with one
intercept per level of `gender`.

```{r}
#| label: model2

model2_mlm <- lmer(
  satisfaction ~ 0 + gender + wnc + partner_wnc +
    recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(model2_mlm)
```

::: {.callout-tip}
## Reading the two intercepts

The fixed-effect coefficients now include `gendermale` and
`genderfemale`. These are the absolute means of each group
(adjusted for the other predictors in the model). Compare to the
DGP values: $\alpha_m = 5.00$ and $\alpha_f = 4.85$.
:::

### Model 3: different intercepts, different slopes (full model)

The full model. Two intercepts, separate slopes for each gender.

```{r}
#| label: model3

model3_mlm <- lmer(
  satisfaction ~ 0 + gender + gender:wnc + gender:partner_wnc +
    gender:recovery + gender:partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(model3_mlm)
```

## The two likelihood ratio tests

```{r}
#| label: lrts

cat("Test A: Model 1 vs Model 2 (Equal vs Different Intercepts)\n")
print(anova(model1_mlm, model2_mlm))

cat("\nTest B: Model 2 vs Model 3 (Equal vs Different Slopes)\n")
print(anova(model2_mlm, model3_mlm))
```

::: {.callout-tip}
## Reading the two tests

- **Test A** tests whether the male and female intercepts differ.
  In our data, the *p*-value should be small — the DGP has a
  0.15-point gap in the intercepts.
- **Test B** tests whether the slopes differ by gender. In our
  data, the *p*-value should be large — the DGP has equal slopes
  across gender.

This is the same information you would get from the [SEM wide
three-LRT sequence](sem-wide.html#the-three-nested-lrts), packaged
as a multilevel model.
:::

## The SEM analogue (wide format)

In `lavaan`, the wide-format model with separate intercepts *is*
the two-intercept model. The "trick" is unnecessary because you
specify the two intercepts explicitly with separate labels.

```{r}
#| label: sem-analogue

model_sem_2int <- '
  satisfaction_a ~ a*wnc_a + p*wnc_p + ar*recovery_a +
                   pr*recovery_p + c*has_children + d*dual_earner
  satisfaction_p ~ a*wnc_p + p*wnc_a + ar*recovery_p +
                   pr*recovery_a + c*has_children + d*dual_earner
  satisfaction_a ~ int_m*1
  satisfaction_p ~ int_f*1
  satisfaction_a ~~ var_h*satisfaction_a
  satisfaction_p ~~ var_w*satisfaction_p
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p
  satisfaction_a ~~ res_cov*satisfaction_p
'

fit_sem_2int <- sem(model_sem_2int, data = ddw)
summary(fit_sem_2int, standardized = TRUE, fit.measures = TRUE)
```

The `int_m` and `int_f` parameters in the SEM correspond to the
`gendermale` and `genderfemale` intercepts in the MLM Model 2.

## When to use this specification

::: {.callout-takeaway}
## Key takeaways

- The two-intercept MLM with `0 + gender` is the cleanest
  specification when you want to report the absolute mean of
  each group.
- The three nested LRTs (Models 1 → 2 → 3) test intercept
  equality and slope equality in a single pipeline.
- The wide-format SEM is mathematically equivalent — the
  "trick" is specific to multilevel models in long format.
- In our data, Model 2 is the recommended baseline: intercepts
  differ, slopes do not.
:::

## What to read next

- The [Hahn et al. (2014) replication
  tutorial](../moderated/hahn2014.html) — the moderated APIM, the
  full APIM workflow.
