---
title: "Moderated APIM: Hahn, Binnewies, & Dormann (2014)"
subtitle: "Replicating the children × partner WNC interaction"
---

The headline result of Hahn, Binnewies, & Dormann (2014, *Journal
of Vocational Behavior*) is that the presence of children in a
household moderates the partner crossover effect of work–nonwork
conflict on relationship satisfaction. Couples with children
*buffer* the negative crossover; couples without children show
the full negative effect.

This tutorial replicates that finding with the simulated
`dyad_data` dataset, using both multilevel and SEM specifications.
The DGP has a small but detectable interaction
(`int_child_pwnc = +0.10`) — your likelihood ratio tests should
recover it.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lme4)
library(lmerTest)
library(lavaan)
library(dplyr)
library(interactions)

load("../../../data/dyad_data.RData")

cat("Dyads with children:    ", sum(ddw$has_children == 1), "of", nrow(ddw), "\n")
cat("Dyads without children: ", sum(ddw$has_children == 0), "of", nrow(ddw), "\n")
```

## MLM specification

The baseline APIM (no moderation) is the same as the
[indistinguishable tutorial](../indistinguishable/mlm.html).

```{r}
#| label: mlm-base

model_base <- lmer(
  satisfaction ~ wnc + partner_wnc + recovery + partner_recovery +
    has_children + dual_earner + (1 | dyad_id),
  data = ddl
)

summary(model_base)
```

The moderated model adds the two interaction terms: partner WNC ×
children, and actor WNC × children.

```{r}
#| label: mlm-moderated

model_mod <- lmer(
  satisfaction ~ wnc + partner_wnc + recovery + partner_recovery +
    has_children + dual_earner +
    partner_wnc:has_children + wnc:has_children +
    (1 | dyad_id),
  data = ddl
)

summary(model_mod)
```

### Likelihood ratio test

```{r}
#| label: mlm-lrt

lrt <- anova(model_base, model_mod)
print(lrt)
```

::: {.callout-tip}
## Reading the LRT
The *p*-value tests the joint null that both interaction terms
(`partner_wnc:has_children` and `wnc:has_children`) are zero. In
our DGP, only the partner-wnc interaction is non-zero (the
actor-wnc interaction is zero), so the LRT is dominated by the
partner-wnc effect.
:::

### Simple slopes

The `interactions` package extracts the partner-WNC effect at each
level of the moderator. With a binary moderator (`has_children`),
this is a two-line output: the slope among dyads *with* children
and the slope among dyads *without* children.

```{r}
#| label: simple-slopes

sim_pwnc <- sim_slopes(model_mod, pred = partner_wnc,
                       modx = has_children, johnson_neyman = FALSE)
print(sim_pwnc)
```

::: {.callout-takeaway}
## Substantive interpretation

Among dyads *without* children, the partner-WNC slope is
`partner_wnc + 0 * partner_wnc:has_children ≈ -0.15`. Among dyads
*with* children, the slope is `partner_wnc + 1 *
partner_wnc:has_children ≈ -0.05`. The negative partner crossover
effect is *attenuated* (buffered) for couples with children —
consistent with Hahn et al. (2014).
:::

### Visualise the interaction

```{r}
#| label: plot-partner
#| fig-cap: "Partner WNC effect on satisfaction, by presence of children. The negative slope is flatter for couples with children."
#| fig-width: 7
#| fig-height: 4

interact_plot(model_mod, pred = partner_wnc,
              modx = has_children, plot.points = TRUE)
```

## SEM specification (wide format)

The SEM analogue creates manual product indicators in the wide
data and fits a wide-format model with the products included.

```{r}
#| label: sem-prep

ddw$wnc_a_children <- ddw$wnc_a * ddw$has_children
ddw$wnc_p_children <- ddw$wnc_p * ddw$has_children
```

```{r}
#| label: sem-moderated

model_sem_mod <- '
  # Actor (male) paths
  satisfaction_a ~ a*wnc_a + p*wnc_p + ar*recovery_a + pr*recovery_p +
    c*has_children + d*dual_earner +
    a_int*wnc_a_children + p_int*wnc_p_children

  # Partner (female) paths
  satisfaction_p ~ a*wnc_p + p*wnc_a + ar*recovery_p + pr*recovery_a +
    c*has_children + d*dual_earner +
    a_int*wnc_p_children + p_int*wnc_a_children

  # Intercepts (equal)
  satisfaction_a ~ int*1
  satisfaction_p ~ int*1

  # Residual variances (equal)
  satisfaction_a ~~ var*satisfaction_a
  satisfaction_p ~~ var*satisfaction_p

  # Predictor variances and covariances
  wnc_a ~~ wnc_p
  recovery_a ~~ recovery_p

  # Residual covariance
  satisfaction_a ~~ res_cov*satisfaction_p

  # --------------------------------------------------------
  # SIMPLE SLOPES (Calculated Parameters)
  # --------------------------------------------------------
  actor_slope_no_child   := a + a_int * 0
  actor_slope_has_child  := a + a_int * 1
  partner_slope_no_child  := p + p_int * 0
  partner_slope_has_child := p + p_int * 1
'

fit_sem_mod <- sem(model_sem_mod, data = ddw)
summary(fit_sem_mod, standardized = TRUE, fit.measures = TRUE)
```

## Compare MLM and SEM

The MLM and SEM should agree on the interaction coefficients
(`p_int` in SEM ≈ `partner_wnc:has_children` in MLM). The
differences will be in the standard errors (Wald in SEM vs.
profile in MLM).

```{r}
#| label: compare
#| message: false
#| warning: false

# MLM partner-wnc × children
cat("MLM partner_wnc:has_children:\n")
cat("  estimate =", round(fixef(model_mod)["partner_wnc:has_children"], 4), "\n")

# SEM p_int
pe <- parameterEstimates(fit_sem_mod)
sem_p_int <- pe[pe$label == "p_int", "est"]
cat("SEM p_int:\n")
cat("  estimate =", round(sem_p_int, 4), "\n")
```

## What to take away

::: {.callout-takeaway}
## Key takeaways

- A moderated APIM tests whether a dyad-level variable shapes
  actor or partner effects. The most common form: does the
  partner crossover effect vary by some dyad-level moderator?
- In `lme4`, the moderation is a direct interaction term in the
  `lmer` formula. In `lavaan`, it is a manual product indicator
  in the wide-format data.
- The two approaches give nearly identical point estimates.
- Simple slopes and visualisations are the recommended way to
  report the conditional effects.
:::

## What to read next

- The [exercises section](../../exercises/index.html) has a
  nine-section problem set on a separate dataset with three
  outcomes and three moderators.
- The [k-patterns page](../../foundations/kpatterns.html) explains
  why the simple-slope approach is more powerful than testing the
  partner effect in isolation.
